---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})
```
# GMMClassifier

The goal of GMMClassifier is to classify single cells (as expression matrix of scRNA-seq data) into subtypes using **G**aussian **M**ixture **M**odel, and to find out the mapping relationship among subtypes across datasets (of different batch).

## Installation

You can install the latest version of GMMClassifier with:

```r
install.packages("devtools")
devtools::install_github("GMMClassifier")
```

## About example data

The example data used in this project is part of [GSE81861](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE81861) of [this paper](https://www.nature.com/articles/ng.3818#accessions). First, the FPKM matrix file of all cells was downloaded:

```bash
aria2c ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE81nnn/GSE81861/suppl/GSE81861_Cell_Line_FPKM.csv.gz
pigz -d GSE81861_Cell_Line_FPKM.csv.gz
```

The matrix was read into R and the columns of two batch were selected as subsets  according to [the results part of the paper](https://www.nature.com/articles/ng.3818#results) "To assess batch effects, we performed scRNA–seq in two batches for GM12878 (lymphoblastoid) cells and also for H1 embryonic stem cells. Gene expression was quantified as fragments per kilobase per million reads (FPKM), and low-quality cells were discarded on the basis of multiple metrics":

```r
library(data.table)
library(tidyverse)
library(org.Hs.eg.db)
library(usethis)
library(genefilter)
library(Biobase)
rowVars <- function(x, ...) {
  rowSums((x - rowMeans(x, ...))^2, ...)/(dim(x)[2] - 1)
}
cl <- fread('GSE81861_Cell_Line_FPKM.csv')
names(cl)[1] <- 'Gene'
cl <- cl[, Ensembl := str_match(Gene, ".+_(.+)\\.\\d+$")[, 2]]
symbols <- mapIds(org.Hs.eg.db, keys = cl$Ensembl, keytype = "ENSEMBL", column="ENTREZID", multiVals = 'first')
cl <- cl[, Entrez := symbols[Ensembl]]
cl <- cl[, c('Gene', 'Ensembl'):= NULL]
cl.b1 <- cbind(cl[, .(Entrez)], cl[, .SD, .SDcols = names(cl) %like% "_B1_"])
cl.b2 <- cbind(cl[, .(Entrez)], cl[, .SD, .SDcols = names(cl) %like% "_B2_"])
cl <- lapply(list(cl.b1, cl.b2), function(x) x[, var := rowVars(.SD), .SDcols = -c('Entrez')])
cl <- lapply(cl, function(x) x[, max.var := max(var), by = 'Entrez'][var != 0 & max.var == var & !is.na(Entrez)])
cl <- lapply(cl, function(x) x[, grep("var", colnames(x)) := NULL])
cl <- lapply(cl, function(x) as.data.frame(x) %>% remove_rownames %>% column_to_rownames(var="Entrez"))
cl <- lapply(cl, function(x) new("ExpressionSet", exprs=as.matrix(x), annotation = 'org.Hs.eg.db'))
cl <- lapply(cl, function(x) nsFilter(x, require.entrez = TRUE, require.GOBP = TRUE, require.GOCC = TRUE, require.GOMF = TRUE, var.cutoff = 0.999))
cl.b1 <- exprs(cl[[1]][[1]])
cl.b2 <- exprs(cl[[2]][[1]])
use_data(cl.b1, compress = 'xz')
use_data(cl.b2, compress = 'xz')
```

Thus, the `matrix` object `cl.b1` contains *GM12878* and *H1* cell of batch 1, and `cl.b2` contains those of batch 2.

## Details for implementation

### How to prepare (lazy-loaded) example data for this R package?

See code in [previous section](#about-example-data). First, all rownames of matrix were converted into Entrez ID, then the features were filtered by [genefilter](https://bioconductor.org/packages/release/bioc/html/genefilter.html) package based on their GO annotation and variance levels. To use fewer features for testing, only features with top 0.1% variance were kepted (about only 14 genes) at the time.

### How to fit a mixture of gaussians to real data?

The [mixtools](https://cran.r-project.org/web/packages/mixtools/index.html) package was used for analyzing mixture models as well as [plotGMM](https://github.com/pdwaggoner/plotGMM) package for visualizing models.

### How can I determine the best number of components when fitting model?

Package mixtools provides a method using [hypothesis testing](https://www.stat.cmu.edu/~cshalizi/402/lectures/20-mixture-examples/lecture-20.pdf) to select number of components. For example:

```{r select_component_num, output.lines = 15, cache=FALSE}
library(mixtools)
library(plotGMM)
data(cl.b1)
a.boot <- boot.comp(y = cl.b1[5, ], max.comp = 10, B = 1000, mix.type = "normalmix", epsilon = 1e-3)
str(a.boot)
```

- `cl.b1[5, ]` means the 5th row for `data.frame` `cl.b1`.
- It seems in rare cases the meaning components(`max.comp`) will exceed `10`.
- The number of bootstrap realization of the likelihood ratio statistic was set to `1000` for more acceptable results.

The `boot.comp()` function performs the likelihood ratio statistic for testing the null hypothesis of a k-component fit versus the alternative hypothesis of a (k+1)-component fit to various mixture models. A p-value is calculated for each test and once the p-value is above a specified significance level, the testing terminates.

> The vertical red lines mark the observed difference in log-likelihoods.

The value of `length(a.boot$obs.log.lik)` was chosen as optimized number of components.

### How to fit a model?

```{r fit_model, cache=FALSE}
# Should not in fast mode: the component densities should be allowed to have different mus and sigmas
a.optim <- normalmixEM(cl.b1[5, ], k = length(a.boot$obs.log.lik))
str(a.optim)
plot_GMM(a.optim, k = length(a.boot$obs.log.lik))
```

Function `normalmixEM` was used to fit a Gaussian Mixture Model. The used number of components was obtained by previous step.

### How to rate components?

To rate a feature, we first rated each component in a model:

- The mean posterior of each component must > 0.9:

```{r check_mean_posterior}
apply(a.optim$posterior, 2, mean) > 0.9
```

- The [overlapping coefficient (OVL)](https://stats.stackexchange.com/q/12216) of the component must < 0.05:

```{r check_OVL}
get_overlap_coef <- function(mu1, mu2, sd1, sd2){
  xs  <- seq(min(mu1 - 4 * sd1, mu2 - 4 * sd2), 
             max(mu1 + 4 * sd1, mu2 + 4 * sd2), 
             length.out = 500)
  f1  <- dnorm(xs, mean = mu1, sd = sd1)
  f2  <- dnorm(xs, mean = mu2, sd = sd2)
  int <- xs[which.max(pmin(f1, f2))]
  l   <- pnorm(int, mu1, sd1, lower.tail = mu1 > mu2)
  r   <- pnorm(int, mu2, sd2, lower.tail = mu1 < mu2)
  l+r
}

get_ovl_iterly <- function(x) {
  # Use nested lapply for calculating one with the others
  ovl <- lapply(seq_len(length(x[['mu']])), function(a) lapply(seq_len(length(x[['mu']])), function(b) ifelse(a == b, NA, get_overlap_coef(x[['mu']][a], x[['mu']][b], x[['sigma']][a], x[['sigma']][b]))))
  (ovl <- lapply(lapply(ovl, unlist), function(x) x[!is.na(x)]))
}

(ovls <- get_ovl_iterly(a.optim))

lapply(ovls, function(x) all(x < 0.05))
```

For instance, the return value of function `get_ovl_iterly()` above is a list contains three elements, each of which is OVLs between one component and others. Finally, each component was tested if all OVLs were smaller than 0.05.

### How to get the (absolute) size of components?

Mixture model provides only ["soft-labels"](http://tinyheero.github.io/2015/10/13/mixture-model.html) (quite different from algorithm like k-means who brings "hard-labels") for samples, which means it need "threshold" for assigning sample to components. Here each of samples was assigned to the component with the biggest posterior probability:

```{r absolute_size}
ab.size <- vector('integer', length = ncol(a.optim$posterior))
names(ab.size) <- colnames(a.optim$posterior)
tmp.size <- table(apply(a.optim$posterior, 1, function(x) colnames(a.optim$posterior)[which.max(x)]))
# Must use <<- here: both <- and assign() do NOT work due to scope
invisible(lapply(names(tmp.size), function(x) ab.size[[x]] <<- tmp.size[x]))
ab.size
```

## Questions

1. What's the difference between "**totally relative percentage**" and "**currently relative percentage**" when rating components?

```{r relative_percentage}
# We have lambda as relative percentage of each components
# But I don't know if the difference
a.optim$lambda
```

2. Which method should be used to (or be implemented as **alternative methods** can be chosen) filter features before fitting the model? One could be filtering based on information of Gene Ontology and variance levels as previous description of data preparation, and others?

3. **Which threshold values** during rating should be implemented as parameters that can be set by user?

- Mean posterior: > 0.9
- Overlapping coefficient: < 0.05
- Absolute size: > 20 cells (of 1966 cells, isn't 1%?)
- Totally relative percentage: 0.005 ~ 0.995
- Currently relative percentage: 0.05 ~ 0.95

3. Should some of these threshold values be tuned by **machine learning** but **not fixed values**? If so, which method should be used?

## TODO list (sorted by priority in descending order)

- [ ] Finish the function for ranking features with rated components.
- [ ] Finish the main function with parameters.
- [ ] Design a method for visualizing the final pseudo decision tree.
- [ ] The result of fitting is somewhat random at the time. Try to make it reproducible when needed.
- [ ] Hide the plotting result of `boot.comp()`. Only keep it in test mode.
